Modularizing Street Map Generation \\ \vspace{8pt} \large An Adventure in Tensors

### 2013-12-12 paper/paper.tex
made json input figure more readable
removed some figure outlines

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}

%borders
\usepackage{float}
\floatstyle{boxed}
\restylefloat{figure}

@@

\begin{abstract}
This paper addresses the issue of creating modular, extensible code given an
existing solution to a problem, namely, interactively modeling large street
networks. We describe a functional implementation of a framework that allows
users to create a realistic street network over an empty domain, based on the
techniques described in \textit{Interactive Procedural Street Modelling}
\cite{chen}. In particular, we were able to express the underlying mathematics
more intuitively and clearly than is possible in an imperative language.
\end{abstract}

\section{Introduction}
Visual artists often face the time-consuming task of generating extensive,
realistic scenes by hand for movies and video games. However, the apparent
randomness found in systems like street layouts can be modelled effectively
with a variety of mathematical algorithms, giving rise to the growing field of
procedural generation. In this paper we will discuss an algorithm using tensor
fields, from \cite{chen}, to generate realistic city street maps. To facilitate
the generation of aesthetically pleasing maps which fit an artist's
requirements, we implement a simple system of user-specified constraints to
allow the user to shape the resulting street map to fit their needs.

@@

While all of the information presented in the paper completely describes all of
the necessary mathematics in a clear manner,
we found the provided implementation of the described
algorithm more difficult to grasp. The large quantity of \cpp code necessary to
implement the full-featured interactive system provided by Chen et al. made it
difficult for us to thoroughly understand the constituent components such that
we would be able to modify or extend the existing code.

\section{Our Goals}
We thought we could use the power and expressiveness of a functional language
to produce a more modular, extensible, and readable implementation of the core
of the algorithm described in Chen et al. which would provide a more
accessible starting point for later work to build off of. While we sacrifice
the performance offered by the \cpp implementation, which is closely tied to
OpenGL, we contend that the gains in programmer productivity will more than
make up for this deficit. While a production implementation of such a pipeline
should certainly be implemented with efficiency in mind, we believe that, in
the process of investigating new algorithms, the ease with which the
implementor can translate an idea or mathematical concept into proof-of-concept
code is of the utmost importance. Later in the paper, we will assess our
success in achieving these goals.

\section{Mathematical Background}
\label{sec:math}
A tensor $t$ is a geometric object represented by a 2x2 symmetric, traceless
matrix:
\[
    \sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
    = \sqmat{a}{b}{b}{-a}
\]
where $R$ is the magnitude and $\theta$ is the direction, and
$R\geq0 \textrm{ and } \theta\in[0,2\pi)$. Since the matrix is symmetric and
traceless, it is only necessary to store two values, $a$ and $b$ to encode
the tensor.

The major eigenvectors of $t$ are
\[
    \left\{
        \lambda\mkvec{\cos{\theta}}{\sin{\theta}} ~|~ \lambda \neq 0
    \right\}
\]

The minor eigenvectors are
\[
    \left\{
        \lambda\mkvec
                {\cos{(\theta+\frac{\pi}{2})}}
                {\sin{(\theta+\frac{\pi}{2})}}
        ~|~ \lambda \neq 0
    \right\}
\]
A tensor field $T$ describes the flow of the constraints across a field,
and is represented as a function that maps from 2D points to tensors. In the
context of this project, a tensor is simply a tool we use to determine the
direction of a field at a point. This direction is found by calculating the
eigenvectors of the tensors. Major eigenvectors are parallel to the field;
minor eigenvectors are perpendicular.

Another important concept is the hyperstreamline. This is an effective way to
visualize tensor fields. It is created by drawing a line through an eigenvector
field such that the line is tangent to every eigenvector along its path. In
this setting, there are major and minor hyperstreamlines that correspond to the
major and minor eigenvectors. The major and minor eigenvectors of a tensor are
guaranteed to be perpendicular for any tensor, unless it is a degenerate point
in the tensor field ($T(\bf p)=0$), for which eigenvectors cannot be
calculated. This property leads to the generation of perpendicular street
intersections, a common feature of real-world road networks, especially in
cities.

\section{Pipeline}
We begin with a set of user constraints, which illustrate the general shape of
the roads on the blank canvas. These can be linear or radial, and can be input
as JSON. From these we generate a tensor field, and from that, major and minor
eigenvector fields, which describe the direction stipulate by the constraints at
every point in the area. Finally, based on these eigenvector fields, we generate
seed points from which to trace streamlines across the map. These streamlines
produce the final roads.

@@

% \floatstyle{plain}
% \restylefloat{figure}
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[b]{2.4in}
        \begin{verbatim}
            [{
              "tycon" : "Linear",
              "posx" : 2,
              "posy" : 2,
              "dir" : 0.01,
              "mag" : 3
            },
            {
              "tycon" : "Radial",
              "posx" : 10,
              "posy" : 10,
              "dir" : 0,
              "mag" : 0
            }]
        \end{verbatim}
        \caption{JSON file from by user}
    \end{subfigure}
    \hspace{20pt}
    \begin{subfigure}[b]{2.4in}
        \todo{image from constraints to left}
        \caption{generated constraints}
    \end{subfigure}
\caption{JSON parsing of constrains}
\label{fig:json}
\end{figure*}
% \floatstyle{boxed}
% \restylefloat{figure}

@@

\paragraph{Linear Constraints}
A linear constraint produces the grid pattern that is a common basis for many
cities (consider the streets of Manhattan). The constraint has two parts: a
magnitude $l$ and a direction $\theta$. The basis field for a linear constraint
is defined as
\[
    T(\pt) =
        l\sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
\]
for any point $\bf p$ in the field. It is apparent from this definition that
the direction of a linear basis field is constant, regardless of the value of
$\bf p$.

@@

\subsection{Eigenvectors}
The eigenvalues of a tensor $t$ represented by the matrix $\sqmat{a}{b}{b}{-a}$
are:
\[
    \lambda = \pm\sqrt{a^2+b^2} \\
\]
The major eigenvalue, $\lambda_1$, is the larger value, and the minor
eigenvalue, $\lambda_2$ is the smaller value. The corresponding eigenvectors
are then
\[
    \epsilon_i = \left\{1,\frac{\lambda_i-a}{b}\right\}
\]
The eigenvectors are defined if the tensor field is not degenerate (that is,
$a$ and $b$ are both zero) at that point \cite{find-evs}.

\subsection{Streamline Tracing}
Mathematically, the basic streamline tracing algorithm is fairly simple,
however, in order to produce a set of streamlines that approximate a road map,
there are several important cases to consider. In order to trace the
streamlines, we are forced to use a heuristic, since it isn’t possible to trace
a continuous line over a constantly-changing tensor field. In order to
approximate this, we choose a step size $d_\textrm{step}$. Beginning at a seed
point, which is chosen using the methods detailed in Section \ref{seedpoints},
we calculate the eigenvector at that point, extend a line in that direction
with length $d_\textrm{step}$, and check if we have reached an end condition.
This process is repeated until the line until one of three end conditions is
reached:
\begin{enumerate}
    \item The point is outside the bounds of the map area
    \item The point is within a certain distance of an existing streamline.
    \item The point is part of a cycle
\end{enumerate}

The first is self-explanatory; if the streamline leaves the boundary of the
map, then we don’t need to draw it any more. The second is one of the
alterations we used, based on the work of \cite{chen}. When designing a set of
roads, it is rare that two streets will approach each other and not connect,
whether they approach at a larger angle (an intersection) or are running mostly
parallel to each other (a merge zone). To approximate this, at each point we
find the nearest neighbor out of the streamlines that have already been drawn.
If it is within a certain separation distance, we draw a line directly between
the two points.

\begin{figure*}[t!]
    \todo{image of that nice circular constraint that worked really well with
    this}
% \label{fig:cycles} name?
\end{figure*}

@@

\floatstyle{plain}
\restylefloat{figure}
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[b]{2.4in}
        \fbox{\includegraphics[width=\textwidth]{images/cycle_nodetection.pdf}}
        \caption{without cycle detection}
    \end{subfigure}
    \hspace{20pt}
    \begin{subfigure}[b]{2.4in}
        \fbox{\includegraphics[width=\textwidth]{images/cycle_detection.pdf}}
        \caption{with cycle detection}
    \end{subfigure}
\caption{Streamline tracing cycle detection}
\label{fig:cycles}
\end{figure*}
\floatstyle{boxed}
\restylefloat{figure}

\subsection{Seed Placement}\label{seedpoints}
To begin constructing a streamline, one must choose a seed point—an initial
coordinate in 2D space from which the streamline is traced. In practice, the
matter of carefully choosing seed points is very important in generating an
evenly spaced configuration of roads. We investigated three methods of placing
seed points:
\begin{enumerate}
    \item Random
    \item Furthest-Point
    \item Even Streamline Spacing
\end{enumerate}
Purely random placement of seed points generally results in both clusters of
roads which are too close together and large gaps between roads. In the
furthest-point placement, we place each new seed point within the bounds of the
space such that the distance to all existing streamlines is maximized. As a
simplifying heuristic for this method, we generate a number of random candidate
seed points which is small compared to the infinite number of possible seed
points (constrained only by floating point precision) and choose the one
furthest from all existing streamlines. In practice, this offers a significant
improvement over the random seed placement, and results in a more even
distribution of streamlines.

@@

We will choose the seed point whose corresponding streamline, when added to the
existing set of streamlines, optimizes the evenness of streamline placement,
according to the above metric. As with the previous method, we employ a
simplifying heuristic, generating a relatively small set of candidate seed
points, tracing a streamline from each, determining which streamline results in
a more even field of streamlines, and adding it.

@@

\begin{figure*}[t!]
\begin{verbatim}
-- returns a tuple of the (larger, smaller) eigenvectors of a tensor
eigenvalues :: Tensor -> (Float, Float)
eigenvalues (Tensor a b d) =
  let eval = sqrt (a*a + b*b)
  in  (eval, (-1)*eval)

-- returns a tuple of the (major, minor) eigenvectors for a tensor
eigenvectors :: Tensor -> (V2.Vector2, V2.Vector2)
eigenvectors (Tensor a 0.0 d) =
  if a > d then (V2.Vector2 1 0, V2.Vector2 0 1)
     else (V2.Vector2 0 1, V2.Vector2 1 0)
     eigenvectors (Tensor a b d) =
       let (eval1, eval2) = eigenvalues (Tensor a b d)
           mkEvec eval    = V2.Vector2 1 ((eval - a) / b)
       in (mkEvec eval1, mkEvec eval2)
\end{verbatim}
\caption{Haskell implementation of eigenvector calculation (from
            \texttt{Tensor.hs})}
\label{fig:evecs}
\end{figure*}

>>>>

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{fancyvrb}

@@

\begin{abstract}
This paper addresses the issue of creating modular, extensible code given an
existing solution to a problem, namely, procedurally modeling large street
networks. We describe a functional implementation of a framework that allows
users to create a realistic street network over an empty domain, based on the
techniques described in \textit{Interactive Procedural Street Modelling}
\cite{chen}. In particular, we were able to express the underlying mathematics
more intuitively and clearly than is possible in an imperative language.
\end{abstract}

\section{Introduction}
Visual artists often face the time-consuming task of generating extensive,
realistic scenes by hand for movies and video games. However, the apparent
randomness found in systems like street layouts can be modelled effectively
with a variety of mathematical algorithms, giving rise to the growing field of
procedural generation. In this paper we will discuss an algorithm using tensor
fields, from \cite{chen}, to generate realistic city street maps. To facilitate
the generation of aesthetically pleasing maps which fit an artist's
requirements, we implement a simple system by which the user can input
constraints to the algorithm, shaping the resulting street map to fit their
needs.

@@

While all of the information presented in the paper completely describes all of
the necessary mathematics for the algorithm in a clear manner, we found the
provided implementation more difficult to grasp. The large quantity of \cpp
code necessary to implement the full-featured interactive system provided by
Chen et al. made it difficult for us to thoroughly understand the constituent
components such that we would be able to modify or extend the existing code.

\section{Our Goals}
We approached the problem with the idea that we could use the power and
expressiveness of a functional language to produce a more modular, extensible,
and readable implementation of the core of the algorithm described in Chen et
al. which would provide a more accessible starting point for later work to
build off of. While we sacrifice the performance offered by the \cpp
implementation, which is closely tied to OpenGL, we contend that the gains in
programmer productivity will alleviate this deficit. While a production
implementation of such a pipeline should certainly be implemented with
efficiency in mind, we believe that, in the process of investigating new
algorithms, the ease with which the implementor can translate an idea or
mathematical concept into proof-of-concept code is of the utmost importance.
Later in the paper, we will assess our success in achieving these goals.

\section{Mathematical Background}\label{sec:math}
In this section we will introduce the mathematical concepts necessary to
understand our algorithm. A tensor $t$ is a geometric object represented by a
2x2 symmetric, traceless matrix:
\[
    \sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
    = \sqmat{a}{b}{b}{-a}
\]
where $R$ is the magnitude and $\theta$ is the direction, and
$R\geq0 \textrm{ and } \theta\in[0,2\pi)$. Since the matrix is symmetric and
traceless, it is only necessary to store two values, $a$ and $b$ to encode
the tensor.

The major eigenvectors of $t$ are:
\[
    \left\{
        \lambda\mkvec{\cos{\theta}}{\sin{\theta}} ~|~ \lambda \neq 0
    \right\}
\]

The minor eigenvectors are:
\[
    \left\{
        \lambda\mkvec
                {\cos{(\theta+\frac{\pi}{2})}}
                {\sin{(\theta+\frac{\pi}{2})}}
        ~|~ \lambda \neq 0
    \right\}
\]

A tensor field $T$ describes the flow of the constraints across a field, and is
represented as a function that maps from 2D points to tensors. In the context
of this project, a tensor a tool we use to determine the direction of a field
at a given point. This direction is found by calculating the eigenvectors of
the tensors. Major eigenvectors are parallel to the field; minor eigenvectors
are perpendicular.

Another important concept is the hyperstreamline, an effective way to visualize
tensor fields. A hyperstreamline is created by drawing a line through an
eigenvector field such that the line is tangent to every eigenvector along its
path. In this setting, there are major and minor hyperstreamlines that
correspond to the major and minor eigenvectors. The major and minor
eigenvectors of a tensor are guaranteed to be perpendicular for any tensor,
unless it is a degenerate point in the tensor field ($T(\bf p)=0$), for which
eigenvectors cannot be calculated. This property leads to the generation of
perpendicular street intersections, a common feature of real-world road
networks, especially in cities.

\section{Pipeline}
We begin with a set of user constraints, which illustrate the general shape of
the roads on the blank canvas. These can be linear or radial, and can be input
as JSON. From these we generate a tensor field, and from that, major and minor
eigenvector fields, which describe the direction stipulated by the constraints
at every point in the area. Finally, based on these eigenvector fields, we
generate seed points from which streamlines are traced across the map. These
streamlines produce the final roads.

@@

% \floatstyle{plain}
% \restylefloat{figure}
\begin{figure*}[t!]
    \centering
    \vspace{10pt}
    \begin{subfigure}[b]{2.4in}
        \begin{verbatim}
            [{
              "tycon" : "Linear",
              "posx" : 2,
              "posy" : 2,
              "dir" : 0.01,
              "mag" : 3
            },
            {
              "tycon" : "Radial",
              "posx" : 10,
              "posy" : 10,
              "dir" : 0,
              "mag" : 0
            }]
        \end{verbatim}
        \caption{JSON file from by user}
    \end{subfigure}
    \hspace{20pt}
    \begin{subfigure}[b]{2.4in}
        \fbox{\includegraphics[width=\textwidth]{images/json.pdf}}
        \vspace{3pt}
        \caption{generated constraints}
    \end{subfigure}
    \caption{JSON input constraint format}
\label{fig:json}
\end{figure*}

@@

\paragraph{Linear Constraints}
A linear constraint produces the grid pattern that is a common basis for many
cities (consider the streets of Manhattan). The constraint has two parts: a
magnitude $l$ and a direction $\theta$. The basis field for a linear constraint
is defined as:
\[
    T(\pt) =
        l\sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
\]
for any point $\bf p$ in the field. It is apparent from this definition that
the direction of a linear basis field is constant, regardless of the value of
$\bf p$.

@@

\subsection{Eigenvectors}
The eigenvalues of a tensor $t$ represented by the matrix $\sqmat{a}{b}{b}{-a}$
are:
\begin{align}
    \lambda_1 &= \sqrt{a^2+b^2} \\
    \lambda_2 &= -\lambda_1 \\
\end{align}

The corresponding eigenvectors are:
\[
    \epsilon_i = \left\{1,\frac{\lambda_i-a}{b}\right\}
\]

The eigenvectors are defined if the tensor field is not degenerate (that is,
$a$ and $b$ are both zero) at that point \cite{find-evs}.

\subsection{Streamline Tracing}
Mathematically, the basic streamline tracing algorithm is fairly simple,
however, in order to produce a set of streamlines that approximate a road map,
there are several important cases to consider. In order to trace the
streamlines, we are forced to use a heuristic, since it isn’t possible to trace
a continuous line over a constantly-changing tensor field. In order to
approximate this, we choose a step size $d_\textrm{step}$. Beginning at a seed
point, which is chosen using the methods detailed in Section \ref{seedpoints},
we calculate the eigenvector at that point, extend a line in that direction
with length $d_\textrm{step}$, and check if we have reached an end condition.
This process is repeated on the line until one of three end conditions is
reached:
\begin{enumerate}
    \item The point is outside the bounds of the map area
    \item The point is within a certain distance of an existing streamline.
    \item The point is part of a cycle
\end{enumerate}

The first is self-explanatory; if the streamline leaves the boundary of the
map, then we don’t need to draw it any more.

The second is one of the alterations we used, based on the work of \cite{chen}.
When designing a set of roads, it is rare that two streets will approach each
other and not connect, whether they approach at a larger angle (an
intersection) or are running mostly parallel to each other (a merge zone). To
approximate this, at each point we find the nearest neighbor out of the
streamlines that have already been drawn.  If it is within a certain separation
distance, we draw a line directly between the two points.

@@

\begin{figure*}[t!]
    \centering
    \fbox{\includegraphics[width=4in]{images/linear.pdf}}
    \caption{Streamline tracing near road merging}
\end{figure*}

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[b]{2.4in}
        \fbox{\includegraphics[width=\textwidth]{images/cycle_nodetection.pdf}}
        \caption{without cycle detection}
    \end{subfigure}
    \hspace{20pt}
    \begin{subfigure}[b]{2.4in}
        \fbox{\includegraphics[width=\textwidth]{images/cycle_detection.pdf}}
        \caption{with cycle detection}
    \end{subfigure}
\caption{Streamline tracing cycle detection}
\label{fig:cycles}
\end{figure*}

\subsection{Seed Placement}\label{seedpoints}
To begin constructing a streamline, one must choose a seed point—--an initial
coordinate in 2D space from which the streamline is traced. In practice, the
matter of carefully choosing seed points is very important in generating an
evenly spaced configuration of roads. We investigated three methods of placing
seed points:
\begin{enumerate}
    \item Random
    \item Furthest-Point
    \item Even Streamline Spacing ($\chi^2$ distribution)
\end{enumerate}

Purely random placement of seed points generally results in both clusters of
roads which are too close together and large gaps between roads.

In the furthest-point placement, we place each new seed point within the bounds
of the space such that the distance to all existing streamlines is maximized.
As a simplifying heuristic for this method, we generate a number of random
candidate seed points which is small compared to the infinite number of
possible seed points (constrained only by floating point precision) and choose
the one furthest from all existing streamlines. In practice, this offers a
significant improvement over the random seed placement, and results in a more
even distribution of streamlines.

@@

We will choose the seed point whose corresponding streamline, when added to the
existing set of streamlines, optimizes the evenness of streamline placement,
according to the above metric. As with the previous method, we employ a
simplifying heuristic, generating a relatively small set of candidate seed
points, tracing a streamline from each, determining which streamline results in
a more even field of streamlines, and adding it to the existing set of
streamlines.

@@

\begin{figure*}[t!]
\begin{Verbatim}[frame=single]
-- returns a tuple of the (larger, smaller) eigenvectors of a tensor
eigenvalues :: Tensor -> (Float, Float)
eigenvalues (Tensor a b d) =
let eval = sqrt (a*a + b*b)
in  (eval, (-1)*eval)

-- returns a tuple of the (major, minor) eigenvectors for a tensor
eigenvectors :: Tensor -> (V2.Vector2, V2.Vector2)
eigenvectors (Tensor a 0.0 d) =
if a > d then (V2.Vector2 1 0, V2.Vector2 0 1)
   else (V2.Vector2 0 1, V2.Vector2 1 0)
   eigenvectors (Tensor a b d) =
     let (eval1, eval2) = eigenvalues (Tensor a b d)
         mkEvec eval    = V2.Vector2 1 ((eval - a) / b)
     in (mkEvec eval1, mkEvec eval2)
\end{Verbatim}

\caption{Haskell implementation of eigenvector calculation (from
            \texttt{Tensor.hs})}
\label{fig:evecs}
\end{figure*}
